---
title: 'Week 2: Examples for Data Tidying'
---


# Revisit Population data obtained from http://transmonee.org/database/

# Visit at the web site of Monitoring the situation of 
# children and women in Europe and Central Asia
# at http://transmonee.org/
# Go to http://transmonee.org/database/ and 
# Download Excel file named
# "Population at the beginning of the year by sex and selected age groups"
# into to your working directory and rename it as Population-1989-2015
# since the file name is very long.


```{r}
library(readxl)
pop_red <- read_excel("data/Population-1989-2015.xlsx", range = cell_rows(5:39))
View(pop_red)
```

# Column 1 and 2 does not have names.

# Investigate the data

```{r}
str(pop_red)
```

# Did you notice any problem?

# The variables 1989-1999 and 2008-2015 are character variables
# What is the reason for it? 
# Actually they should have been numeric!..

# There are empty rows in the sheet and they are denoted by NA in data viewer panel
# There are missing values in the sheet and they are denoted by "-" in the sheet as well as in the data
# viewer panel.
# We do not need the second column is not needed.

# First drop all the rows which are completely missing

# drop_na() drops rows who are missing at all columns--complete case analysis

```{r}
pop_red_tidy <- pop_red %>%
                 drop_na() %>% 
View(pop_red_tidy)
```

```{r}
str(pop_red_tidy)
```


# Rename the first column as "Country", drop the second column, and
# sorty by Country


```{r}
pop_red_tidy_more <- pop_red_tidy %>%
                      drop_na() %>% 
                      rename("Country"="...1") %>%
                      select(-"...2") %>% 
                      arrange(Country) 
            
View(pop_red_tidy_more)
```


# Investigate the data

```{r}
str(pop_red_tidy_more)
```

# We need to replace "-" with "NA".

```{r}
pop_red_withna <- pop_red_tidy_more %>% 
                  na_if("-") #from dplyr

View(pop_red_withna)
```

```{r}
str(pop_red_withna)
```

# Except the Country column, convert character variables
# into numeric variables

```{r}
pop_red_tidy_tidy <- pop_red_withna %>% 
                     mutate_at(vars("1989":"2015"), as.numeric) 
str(pop_red_tidy_tidy)
View(pop_red_tidy_tidy)                  
```                    


# Lastly, turn this data into long format so that
# We can use it later (for visualization etc)


```{r}
pop_red_long <- pop_red_tidy_tidy %>% 
                pivot_longer(!Country, names_to="Year", values_to="Population")
View(pop_red_long)
```

