---
title: 'Week 04: Examples for Data Tidying'
---


# Revisit Population data obtained from http://transmonee.org/database/

# Visit at the web site of Monitoring the situation of 
# children and women in Europe and Central Asia
# at http://transmonee.org/
# Go to http://transmonee.org/database/ and 
# Download Excel file named
# "Population at the beginning of the year by sex and selected age groups"
# into to your working directory and rename it as Population-1989-2015
# since the file name is very long.


```{r}
library(readxl)
pop_red <- read_excel("data/Population-1989-2019.xlsx", range = cell_rows(5:40))
View(pop_red)
```

# Column 1 and 2 does not have names.

# Investigate the data

```{r}
str(pop_red)
```

# Did you notice any problem?

# The variables 1989-2019 are character variables
# What is the reason for it? 
# Actually they should have been numeric!..

# There are empty rows in the sheet and they are denoted by NA in data viewer panel
# There are missing values in the sheet and they are denoted by "-" in the sheet as well as in the data
# viewer panel.
# We do not need the second column.

# First drop all the rows which are "completely missing".

# drop_na() drops rows which are missing at all columns--complete case analysis

```{r}
library(tidyr)
pop_red_tidy <- pop_red %>%
                 drop_na() 
View(pop_red_tidy)
```

```{r}
str(pop_red_tidy)
```


# Rename the first column as "Country", drop the second column, and
# sort by Country


```{r}
pop_red_tidy_more <- pop_red_tidy %>%
                      drop_na() %>% 
                      rename(Country = "...1") %>%
                      select(-"...2") %>% #unquoted ...2 is also OK, but has low readability.
                      arrange(Country) 
            
View(pop_red_tidy_more)
```


# Investigate the data

```{r}
str(pop_red_tidy_more)
```

# We need to replace "-" with "NA".

```{r}
# read na_if() as: if "-", then it is na
pop_red_withna <- pop_red_tidy_more %>% 
                  na_if("-")  #from dplyr 
  
View(pop_red_withna)
```

```{r}
str(pop_red_withna)
```

# Except the Country column, convert character variables
# into numeric variables with mutate_

```{r}
pop_red_tidy_tidy <- pop_red_withna %>% 
                     #mutate_at(vars(1989:2019), as.numeric) #not works
                     mutate_at(vars("1989":"2019"), as.numeric) #not as.numeric(), () drops
str(pop_red_tidy_tidy)
``` 

```{r}

View(pop_red_tidy_tidy)                  
```                    

# Lastly, this data is in wide format.
# Turn this data into long format so that
# we can use it later (for visualization etc)

# except country variable, take the columns, add column
# names (values) to under a new column named "Year", and
# assign the values to another column named "Population".

```{r}
pop_red_long <- pop_red_tidy_tidy %>% 
                pivot_longer(!Country, names_to="Year", values_to="Population")
View(pop_red_long)
```

